Mergeable Types

### 2017-02-26 paper/model.tex
Edits.

\input{list-code}

{\bf Lists}. List, as an abstract data type, supports three
operations: \C{insert x i l}, that inserts an element \C{x} at
position \C{i} in the list \C{l}, \C{delete i l}, that deletes the
element at \C{i}'th position, and \C{subst i x l}, that substitutes
the element at \C{i}'th position with \C{x}. Lists of mergeable items
are mergeable. The signature of a mergeable list library is shown in
Fig.~\ref{fig:mlist-sig}. Note that the type of items in the list
(\C{A.t}) is also required to be mergeable. The signature
\C{MERGEABLE} will be discussed shortly.

The merge operation for lists is composed of two separate functions,
\C{edit\_seq} and \C{op\_transform}.
\begin{itemize}
  \item \C{edit\_seq} takes a pair of lists, \C{v} and $\C{v'}$, and
  computes the shortest sequence of list operations that need to be
  applied on \C{v} to obtain \C{v'}. Such a sequence is called an
  \emph{edit sequence}. The length of the sequence corresponds to the
  standard notion of the \emph{edit distance} between the two lists,
  which can be computed in polynomial time, for e.g., using the
  Wagner-Fisher algorithm~\cite{wagner-fischer}. A slight modification
  of Wagner-Fischer also lets us compute the edit sequence. The
  implementation in Fig.~\ref{fig:mlist} represents edits using the
  type \C{edit}. Constructors \C{I}, \C{D}, and \C{S} stand for
  \C{insert}, \C{delete}, and \C{subst}, respectively; \C{Nop} is
  there for technical reasons. The subst constructor also carries the
  \C{A.t} element that was substituted. Our implementation of
  Wagner-Fischer is more-or-less standard (hence, not shown), except
  that the edit sequence it returns orders \C{I}'s and \C{D}'s before
  \C{S}'s.  Fig.~\ref{fig:list-eg} illustrates edit sequences for a
  sample list. The sequence \C{[I(c,0); S(3,c,s)]} maps the list
  \C{[a;b;c]} to \C{[c;a;b;s]} because \C{subst 3 s (insert c 0
  [a;b;c]) = [c;a;b;s]}.
  \item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and
  $s_2$, that map a list $v$ to two different lists, $v_1$ and $v_2$
  (e.g., Fig.~\ref{fig:list-eg}), and transforms $s_1$ to $s_1'$ such
  that $s_1'$ has the same effect on $v_2$ as $s_1$ had on $v$.  For
  instance, in Fig.~\ref{list-eg}, let $s_1$ be the edit sequence
  \C{[D(1); S(d,1)]}, which maps the list \C{[a;b;c]} ($v$) to
  \C{[a;d]} ($v_1$). The \C{D} edit deletes the first element of $v$.
  However, if $s_1$ is to be applied on the list \C{[c;a;b;s]}
  ($v_2$), which has a new element at position 0, the \C{D} edit must
  delete the the second element from $v_2$, if it were to have the
  same effect as it did on $v$. We say \C{D(1)} in $s_1$ has to be
  transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The
  function \C{op\_transform} of Fig.~\ref{fig:mlist-xform} computes
  such \emph{operational transformation} for the sequence \C{mine}
  w.r.t the sequence \C{others}, and returns \C{mine'} as the
  transformed sequence.  
\end{itemize}

>>>>

\input{list-sig}

{\bf Lists}. List, as an abstract data type, supports three
operations: \C{insert x i l}, to insert an element \C{x} at position
\C{i} in the list \C{l}, \C{delete i l}, to delete the element at
\C{i}'th position, and \C{subst i x l}, to substitute the element at
\C{i}'th position with \C{x}. Lists of mergeable items are mergeable
while preserving the following properties:
\begin{itemize}
  \item Elements present in both the merging lists will be present in
  the final list. No element is deleted unless it is deleted in at
  least one list.
  \item An element deleted in one list will not reappear in the merged
  list.
  \item The order of elements is retained. That is, if $x$ occurs
  before $y$ in one of the lists, and if both elements are present in
  the merged list, then $x$ occurs before $y$ also in the merged list.
  \item If an element $y$ substitutes $x$ in one list, then $y$ also
  substitutes $x$ in the merged list (i.e., $x$ is absent and $y$ is
  present), unless $x$ is deleted or substituted in the other list. In
  the former case, deletion wins, and neither $x$ nor $y$ occur in the
  merged list. In the latter case, if $x$ is substituted by a
  different element $z$ in the other list, then the merged list
  substitutes $x$ by a merge of $y$ and $z$, defined as per the merge
  semantics of the list element type.
\end{itemize}
Thus list is an example of a polymorphic mergeable data type that
demonstrates how mergeability can be composed. The compositionality is
reified in the the signature of a mergeable list library is shown in
Fig.~\ref{fig:mlist-sig}; observe that the type of items in the list
(\C{A.t}) is also required to be mergeable (\C{A.merge}).

The merge operation for lists is composed of two separate functions,
\C{edit\_seq} and \C{op\_transform}. Both these functions have been
implemented in less than 50 lines of OCaml included in the appendix.
The functions are described below.
\begin{itemize}
  \item \C{edit\_seq} takes a pair of lists, \C{v} and $\C{v'}$, and
  computes the shortest sequence of list operations that need to be
  applied on \C{v} to obtain \C{v'}. Such a sequence is called an
  \emph{edit sequence}. The length of the sequence corresponds to the
  standard notion of the \emph{edit distance} between the two lists,
  which can be computed in polynomial time, for e.g., using the
  Wagner-Fisher algorithm~\cite{wagner-fischer}. \C{edit\_seq} uses a
  slight modification of Wagner-Fischer, that also lets it compute the
  edit sequence (an \C{edit list}). The implementation specified in
  Fig.~\ref{fig:mlist-sig} represents edits using the type \C{edit}.
  Constructors \C{I}, \C{D}, and \C{S} stand for \C{insert},
  \C{delete}, and \C{subst}, respectively\footnote{
    Note that \C{I}, \C{D} and \C{S} are \emph{not} effects; no list
    function generates them, and they are not exchanged between
    concurrent threads. They are simply tags used by the list merge
    operation for convenience.
  }. The subst constructor also carries the \C{A.t} element that was
  substituted. The element serves as the \C{lca} argument to
  \C{A.merge} in case of concurrent conflicting substitutions.
  Fig.~\ref{fig:list-eg} illustrates edit sequences for a sample list.
  The sequence \C{[I(c,0); S(3,c,s)]} maps the list \C{[a;b;c]} to
  \C{[c;a;b;s]} because \C{subst 3 s (insert c 0 [a;b;c]) =
  [c;a;b;s]}.
  \item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and
  $s_2$, that map a list $v$ to two different lists, $v_1$ and $v_2$
  (e.g., Fig.~\ref{fig:list-eg}), and transforms $s_1$ to $s_1'$ such
  that $s_1'$ has the same effect on $v_2$ as $s_1$ had on $v$.  For
  instance, in Fig.~\ref{list-eg}, let $s_1$ be the edit sequence
  \C{[D(1); S(1,c,d)]}, which maps the list \C{[a;b;c]} ($v$) to
  \C{[a;d]} ($v_1$). The \C{D} edit deletes the first element of $v$.
  However, if $s_1$ is to be applied on the list \C{[c;a;b;s]}
  ($v_2$), which has a new element at position 0, the \C{D} edit must
  delete the the second element from $v_2$, if it were to have the
  same effect as it did on $v$. We say \C{D(1)} in $s_1$ has to be
  transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The
  function \C{op\_transform} (Fig.~\ref{fig:mlist-sig}) computes such
  \emph{operational transformations} for an edit sequence $s_1$ w.r.t
  another edit sequence $s_2$. A notable transformation rule is for a
  substitution w.r.t another substitution at the same position. Since
  the substituting items could be different, the function relies on
  the item merge function (\C{A.merge}) to merge them into a single
  item. For instance, in Fig.~\ref{fig:list-eg}, there are
  simultaneous substitutions to the item \C{c} in $s_1$ and $s_2$;
  while $s_1$ substitutes it with \C{d}, $s_2$ does it with \C{s}. The
  merged list therefore substitutes \C{c} with the merge of \C{d} and
  \C{s}, defined as per the merge semantics of \C{A.t}.
\end{itemize}
The \C{MList.merge} can now be defined as a function that first
computes edit sequences, $s_1$ and $s_2$, for the two concurrent
lists, $v_1$ and $v_2$ w.r.t their common ancestor $v$, transforms one
of them (say, $s_1$) w.r.t $s_2$ to obtain $s_1'$, and finally applies
(the operations defined by) $s_1'$ to $v_2$ to obtain the merged list.

A signficant advantage of mergeable types is their composability. For
instance, a mergeble list of counters can be obtained simply by
instantiating \C{MList} for \C{Counter} (i.e., \C{MList(Counter)}). A
non-trivial demonstration of the utility of such composability is
given below.

% The mergeable list data type demonstrates a common pattern of writing
% \C{merge} functions applicable to a range of functional data
% structures. For instance, a binary tree (with rotations) can be made
% mergeable in the same way as list, i.e., by computing edit sequences
% (\C{edit\_seq}) and then their transformations
% (\C{op\_transform})\footnote{
%   Like Wagner-Fischer, established algorithms exist for computing tree
%   edit distances~\cite{tree-diff}. An alternative naive method to
%   merge a pair of trees is by merging the lists obtained from their
%   in-order traversal, followed by reconstructing the tree.
% }.

{\bf Shopping List}. A shopping list is an
